Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-09-12 Documents/M1/realizableFermionicM1/sections/imex.tex
Edits throughout the paper due to extensive comments from Cory.

\section{Constraint-Preserving IMEX Schemes}
\label{sec:imex}

In this section we discuss the class of IMEX schemes that are used for the realizability-preserving DG-IMEX method developed in Section~\ref{sec:realizableDGIMEX} (see also \ref{app:butcherTables} for additional details).  
The semi-discretization of the moment equations with the DG method given by Eq.~\eqref{eq:semidiscreteDG} results in a system of ordinary differential equations (ODEs) in each element of the form
\begin{equation}
  \dot{\vect{u}}
  =\vect{\cT}(\vect{u})+\f{1}{\tau}\,\vect{\cQ}(\vect{u}),
  \label{eq:ode}
\end{equation}
where $\vect{u}$ are the degrees of freedom evolved with the DG method; i.e., for a test space spanned by $\{\phi_{i}(\vect{x})\}_{i=1}^{N}\in\bbV^{k}$, we let
\begin{equation}
  \vect{u}=\f{1}{|\bK|}\Big(\,\int_{\bK}\vect{\cM}_{h}\,\phi_{1}\,d\vect{x},\int_{\bK}\vect{\cM}_{h}\,\phi_{2}\,d\vect{x},\ldots,\int_{\bK}\vect{\cM}_{h}\,\phi_{N}\,d\vect{x}\,\Big)^{T}.
\end{equation}
Thus, for $\phi_{1}=1$, the first components of $\vect{u}$ are the cell averaged moments.  
In Eq.~\eqref{eq:ode}, the transport operator $\vect{\cT}$ corresponds to the second and third term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}, while the collision operator $\vect{\cQ}$ corresponds to the right-hand side of Eq.~\eqref{eq:semidiscreteDG}.

Eq.~\eqref{eq:ode} is to be integrated forward in time with an ODE integrator, and we begin with the following definition of a constraint-preserving time integration scheme, as applied to integrate a physical model forward in time.  
\begin{define}
  Given sufficient conditions, a constraint-preserving time integration scheme preserves the constraints of a model if the set of admissible states satisfying the constraints form a convex set.  
  \label{def:constraintPreserving}
\end{define}
As an example, given a spatial discretization of a model that is constraint-preserving with the forward Euler method for a sufficiently small time step, high-order explicit strong stability-preserving Runge-Kutta (SSP-RK) methods, which can be expressed as convex combinations of forward Euler steps, also preserve the constraints (possibly with a smaller time step), and form a class of constraint-preserving schemes.

Since the realizable set $\cR$ is convex, constraint-preserving schemes can be used to design realizability-preserving schemes for the two-moment model.

\subsection{Second-Order Accurate, Constraint-Preserving IMEX Schemes}

In many applications, collisions with a background induce stiffness ($\tau\ll1$) in regions of the computational domain that must be treated with implicit methods.  
Meanwhile, the time scales induced by the transport term can be treated with explicit methods.  
This motivates the use of IMEX methods \cite{ascher_etal_1997,pareschiRusso_2005}.  
Our goal is to employ IMEX schemes that preserve realizability of the moments, subject only to a time step governed by the explicit transport operator and comparable to the time step required for numerical stability of the explicit scheme.  
We seek to achieve this goal with constraint-preserving IMEX schemes.

Unfortunately, high-order (second or higher order temporal accuracy) constraint-preserving IMEX methods with time step restrictions solely due to the transport operator do not exist (see for example Proposition~6.2 in \cite{gottlieb_etal_2001}, which rules out the existence of implicit SSP-RK methods of order higher than one).  
To overcome this barrier, Chertock et al. \cite{chertock_etal_2015} presented IMEX schemes with a correction step.  
These schemes are SSP but only first-order accurate within the standard IMEX framework.  
The correction step is introduced to recover second-order accuracy.  
However, the correction step in \cite{chertock_etal_2015} involves both the transport and collision operators, and we have found that, when applied to the fermionic two-moment model, realizability is subject to a time step restriction that depends on $\tau$, which may become too restrictive for stiff problems.  
More recently, Hu et al. \cite{hu_etal_2018}, presented similar IMEX schemes for problems involving BGK-type collision operators, but with a correction step that does not include the transport operator.  
In this case, the scheme is constraint-preserving, subject only to time step restrictions stemming from the transport operator, which is more attractive for our target application.  
These second-order accurate, $s$-stage IMEX schemes take the following form \cite{hu_etal_2018}
\begin{align}
  \vect{u}^{(i)}
  &=\vect{u}^{n}
  +\dt\sum_{j=1}^{i-1}\tilde{a}_{ij}\,\vect{\cT}(\vect{u}^{(j)})
  +\dt\sum_{j=1}^{i}a_{ij}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(j)}),
  \quad i=1,\ldots,s, \label{imexStages} \\
  \tilde{\vect{u}}^{n+1}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{s}\tilde{w}_{i}\,\vect{\cT}(\vect{u}^{(i)})
  +\dt\sum_{i=1}^{s}w_{i}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}), \label{imexIntermediate} \\
  \vect{u}^{n+1}
  &=\tilde{\vect{u}}^{n+1}-\alpha\,\dt^{2}\,\f{1}{\tau^{2}}\,\vect{\cQ}'(\vect{u}^{*})\,\vect{\cQ}(\vect{u}^{n+1}), \label{eq:imexCorrection}
\end{align}
where, as in standard IMEX schemes, $(\tilde{a}_{ij})$ and $(a_{ij})$, components of $s\times s$ matrices $\tilde{A}$ and $A$, respectively, and the vectors $\tilde{\vect{w}}=(\tilde{w}_{1},\ldots,\tilde{w}_{s})^{T}$ and $\vect{w}=(w_{1},\ldots,w_{s})^{T}$ must satisfy certain order conditions \cite{pareschiRusso_2005}.  
The coefficient $\alpha$ in the correction step is positive, and $\vect{\cQ}'$ is the Fr{\'e}chet derivative of the collision term evaluated at $\vect{u}^{*}$.  
For second-order accuracy, $\vect{\cQ}'$ can be evaluated using any of the stage values ($\vect{u}^{n}$, $\vect{u}^{(i)}$, or $\tilde{\vect{u}}^{n+1}$).  
For second-order temporal accuracy, the order conditions for the IMEX scheme in Eqs.~\eqref{imexStages}-\eqref{eq:imexCorrection} are \cite{hu_etal_2018}
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=\sum_{i=1}^{s}w_{i}=1,
  \label{orderConditions1}
\end{equation}
and
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}\tilde{w}_{i}\,c_{i}
  =\sum_{i=1}^{s}w_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}w_{i}\,c_{i}-\alpha=\f{1}{2}, 
  \label{orderConditions2}
\end{equation}
where $\tilde{c}_{i}$ and $c_{i}$ are given in \ref{app:butcherTables}.  
For globally stiffly accurate (GSA) IMEX schemes, $\tilde{w}_{i}=\tilde{a}_{si}$ and $w_{i}=a_{si}$ for $i=1,\ldots,s$, so that $\tilde{\vect{u}}^{n+1}=\vect{u}^{(s)}$ \cite{ascher_etal_1997}.  
This property is beneficial for very stiff problems, and also simplifies the proof of the realizability-preserving property of the DG-IMEX scheme given in Section~\ref{sec:realizableDGIMEX} since it eliminates the assembly step in Eq.~\eqref{imexIntermediate}.

Hu et al. \cite{hu_etal_2018} rewrite the stage values in Eq.~\eqref{imexStages} in the following form
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$ (see \ref{app:butcherTables}).  
In Eq.~\eqref{eq:imexStagesRewrite}, $\vect{u}^{(0)}=\vect{u}^{n}$.  
Two types of IMEX schemes are considered: \emph{type~A} \cite{pareschiRusso_2005,dimarcoPareschi2013} and \emph{type~ARS} \cite{ascher_etal_1997}.  
For IMEX schemes of type~A, the matrix $A$ is invertible.  
For IMEX schemes of type~ARS, the matrix $A$ can be written as
\begin{equation*}
  \left( 
    \begin{matrix} 
       0 & 0 \\ 
       0 & \hat{A}
    \end{matrix}
  \right),
\end{equation*}
where $\hat{A}$ is invertible.  
In writing the stages in the IMEX scheme in the general form given by Eq.~\eqref{eq:imexStagesRewrite}, it should be noted that $\tilde{c}_{i0}=0$ for IMEX schemes of Type~A, and $c_{i1}=\tilde{c}_{i1}=0$ for IMEX schemes of Type~ARS \cite{hu_etal_2018}.  
Type A schemes can be made constraint-preserving by requiring 
\begin{align}
  &a_{ii}>0, \quad c_{i0}\ge0, \quad \text{for} \quad i=1,\ldots,s, \nonumber \\
  &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=2,\ldots,s, \quad\text{and}\quad j=1,\ldots,s-1.  
  \label{eq:positivityConditionsTypeA}
\end{align}
Similarly, type ARS schemes can be made constraint-preserving by requiring 
\begin{align}
  &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
  &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1.  
  \label{eq:positivityConditionsTypeARS}
\end{align}
Coefficients were given in \cite{hu_etal_2018} for GSA schemes of type A with $s=3$ and type ARS with $s=4$.  
(It was also proven that $s=3$ and $s=4$ are the necessary number of stages needed for GSA second-order constraint-preserving IMEX schemes of type A and type ARS, respectively.)

For the scheme to be constraint-preserving, the coefficients $c_{ij}$ in Eq.~\eqref{eq:imexStagesRewrite} must also satisfy $\sum_{j=0}^{i-1}c_{ij}=1$, which implies the so-called Shu-Osher form of the explicit part \cite{shuOsher_1988}.  
If the expression inside the square brackets in Eq.~\eqref{eq:imexStagesRewrite} --- which is in the form of a forward Euler update with time step $\hat{c}_{ij}\,\dt$ --- is in the (convex) set of admissible states for all $i=1,\ldots,s$, and $j=0,\ldots,i-1$, then it follows from convexity arguments that the entire sum on the right-hand side of Eq.~\eqref{eq:imexStagesRewrite} is also admissible.  
Thus, if the explicit update with the transport operator is admissible for a time step $\dt_{\mbox{\tiny Ex}}$, the IMEX scheme is constraint-preserving for a time step $\dt\le c_{\mbox{\tiny Sch}}\,\dt_{\mbox{\tiny Ex}}$, where
\begin{equation}
  c_{\mbox{\tiny Sch}}=\min_{\substack{i = 2,\ldots,s \\ 
       j = 1,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}} \quad \text{(Type A)}, \quad  c_{\mbox{\tiny Sch}}=\min_{\substack{i = 2,\ldots,s \\ 
              j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}} \quad \text{(Type ARS)}.
  \label{eq:imexCFL}
\end{equation}
Here, $c_{\mbox{\tiny Sch}}$ is the CFL condition, relative to $\dt_{\mbox{\tiny Ex}}$, for the IMEX scheme to be constraint-preserving.  
It is desirable to make $c_{\mbox{\tiny Sch}}$ as large (close to $1$) as possible.  
Note that for $\vect{u}^{(i)}$ to be admissible also requires the implicit solve (equivalent to implicit Euler) to be constraint-preserving.  
In \cite{hu_etal_2018}, Hu et al. provide examples of GSA, constraint-preserving IMEX schemes of type A (see scheme PA2 in \ref{app:butcherTables}) and type ARS.  
In \ref{app:butcherTables}, we provide another example of a GSA, constraint-preserving IMEX scheme of type A (scheme PA2+), with a larger $c_{\mbox{\tiny Sch}}$ (a factor of about $1.7$ larger).

\subsection{Constraint-Preserving, Diffusion Accurate IMEX Schemes}

Unfortunately, the correction step in Eq.~\eqref{eq:imexCorrection} deteriorates the accuracy of the IMEX scheme when applied to the moment equations in the diffusion limit.  
The diffusion limit is characterized by frequent collisions ($\tau\ll 1$) in a purely scattering medium ($\xi=0$), and exhibits long-time behavior governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
Here the time derivative term in the equation for the particle flux has been dropped (formally $\cO(\tau^{2})$) so that, to leading order in $\tau$, the second equation in Eq.~\eqref{eq:diffusionLimit} states a balance between the transport term and the collision term.  
Furthermore, in the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$.  
The absence of the transport operator in the correction step in Eq.~\eqref{eq:imexCorrection}, destroys the balance between the transport term and the collision term.  
We demonstrate the inferior performance of IMEX schemes with this correction step in the diffusion limit in Section~\ref{sec:smoothProblems}.  
We have also implemented and tested one of the IMEX schemes in Chertock et al. \cite{chertock_etal_2015} (not included in Section~\ref{sec:smoothProblems}), where the transport operator is part of the correction step, and found it to perform very well in the diffusion limit.  
However, we have not been able to prove the realizability-preserving property with this approach without invoking a too severe time step restriction.  
We therefore proceed to design constraint-preserving IMEX schemes, without the correction step, that perform better in the diffusion limit.  
We limit the scope to IMEX schemes of type~ARS.  
(It can be shown that IMEX schemes of type~A conforming to Definition~\ref{def:PD-IMEX} below do not exist; cf. \ref{app:noTypeA}.)

>>>>

\section{Convex-Invariant IMEX Schemes}
\label{sec:imex}

In this section we discuss the class of IMEX schemes that are used for the realizability-preserving DG-IMEX method developed in Section~\ref{sec:realizableDGIMEX} (see also \ref{app:butcherTables} for additional details).  
The semi-discretization of the moment equations with the DG method given by Eq.~\eqref{eq:semidiscreteDG} results in a system of ordinary differential equations (ODEs) of the form
\begin{equation}
  \dot{\vect{u}}
  =\vect{\cT}(\vect{u})+\f{1}{\tau}\,\vect{\cQ}(\vect{u}),
  \label{eq:ode}
\end{equation}
where $\vect{u}=\{\vect{u}_{\bK}\}_{\bK\in\mathscr{T}}$ are the degrees of freedom evolved with the DG method; i.e., for a test space spanned by $\{\phi_{i}(\vect{x})\}_{i=1}^{N}\in\bbV^{k}$, we let
\begin{equation}
  \vect{u}_{\bK}=\f{1}{|\bK|}\Big(\,\int_{\bK}\vect{\cM}_{h}\,\phi_{1}\,d\vect{x},\int_{\bK}\vect{\cM}_{h}\,\phi_{2}\,d\vect{x},\ldots,\int_{\bK}\vect{\cM}_{h}\,\phi_{N}\,d\vect{x}\,\Big)^{T}.
\end{equation}
Thus, for $\phi_{1}=1$, the first components of $\vect{u}_{\bK}$ are the cell averaged moments.  
In Eq.~\eqref{eq:ode}, the transport operator $\vect{\cT}$ corresponds to the second and third term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}, while the collision operator $\vect{\cQ}$ corresponds to the right-hand side of Eq.~\eqref{eq:semidiscreteDG}.

Eq.~\eqref{eq:ode} is to be integrated forward in time with an ODE integrator.  
Since the realizable set $\cR$ is convex, convex-invariant schemes can be used to design realizability-preserving schemes for the two-moment model.  
\begin{define}
  Given sufficient conditions, a convex-invariant time integration scheme preserves the constraints of a model if the set of admissible states satisfying the constraints forms a convex set.  
  \label{def:constraintPreserving}
\end{define}
As an example, high-order explicit strong stability-preserving Runge-Kutta (SSP-RK) methods form a class of convex-invariant schemes.

\subsection{Second-Order Accurate, Convex-Invariant IMEX Schemes}

In many applications, collisions with a background induce stiffness ($\tau\ll1$) in regions of the computational domain that must be treated with implicit methods.  
Meanwhile, the time scales induced by the transport term can be treated with explicit methods.  
This motivates the use of IMEX methods \cite{ascher_etal_1997,pareschiRusso_2005}.  
Our goal is to employ IMEX schemes that preserve realizability of the moments, subject only to a time step governed by the explicit transport operator and comparable to the time step required for numerical stability of the explicit scheme.  
We seek to achieve this goal with convex-invariant IMEX schemes.

Unfortunately, high-order (second or higher order temporal accuracy) convex-invariant IMEX methods with time step restrictions solely due to the transport operator do not exist (see for example Proposition~6.2 in \cite{gottlieb_etal_2001}, which rules out the existence of implicit SSP-RK methods of order higher than one).  
To overcome this barrier, Chertock et al. \cite{chertock_etal_2015} presented IMEX schemes with a correction step.  
These schemes are SSP but only first-order accurate within the standard IMEX framework.  
The correction step is introduced to recover second-order accuracy.  
However, the correction step in \cite{chertock_etal_2015} involves both the transport and collision operators, and we have found that, when applied to the fermionic two-moment model, realizability is subject to a time step restriction that depends on $\tau$ in a way that becomes too restrictive for stiff problems.  
More recently, Hu et al. \cite{hu_etal_2018}, presented similar IMEX schemes for problems involving BGK-type collision operators, but with a correction step that does not include the transport operator.  
In this case, the scheme is convex-invariant, subject only to time step restrictions stemming from the transport operator, which is more attractive for our target application.  
These second-order accurate, $s$-stage IMEX schemes take the following form \cite{hu_etal_2018}
\begin{align}
  \vect{u}^{(i)}
  &=\vect{u}^{n}
  +\dt\sum_{j=1}^{i-1}\tilde{a}_{ij}\,\vect{\cT}(\vect{u}^{(j)})
  +\dt\sum_{j=1}^{i}a_{ij}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(j)}),
  \quad i=1,\ldots,s, \label{imexStages} \\
  \tilde{\vect{u}}^{n+1}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{s}\tilde{w}_{i}\,\vect{\cT}(\vect{u}^{(i)})
  +\dt\sum_{i=1}^{s}w_{i}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}), \label{imexIntermediate} \\
  \vect{u}^{n+1}
  &=\tilde{\vect{u}}^{n+1}-\alpha\,\dt^{2}\,\f{1}{\tau^{2}}\,\vect{\cQ}'(\vect{u}^{*})\,\vect{\cQ}(\vect{u}^{n+1}), \label{eq:imexCorrection}
\end{align}
where, as in standard IMEX schemes, $(\tilde{a}_{ij})$ and $(a_{ij})$, components of $s\times s$ matrices $\tilde{A}$ and $A$, respectively, and the vectors $\tilde{\vect{w}}=(\tilde{w}_{1},\ldots,\tilde{w}_{s})^{T}$ and $\vect{w}=(w_{1},\ldots,w_{s})^{T}$ must satisfy certain order conditions \cite{pareschiRusso_2005}.  
The coefficient $\alpha$ in the correction step is positive, and $\vect{\cQ}'$ is the Fr{\'e}chet derivative of the collision term evaluated at $\vect{u}^{*}$.  
For second-order accuracy, $\vect{\cQ}'$ can be evaluated using any of the stage values ($\vect{u}^{n}$, $\vect{u}^{(i)}$, or $\tilde{\vect{u}}^{n+1}$).  
For second-order temporal accuracy, the order conditions for the IMEX scheme in Eqs.~\eqref{imexStages}-\eqref{eq:imexCorrection} are \cite{hu_etal_2018}
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=\sum_{i=1}^{s}w_{i}=1,
  \label{orderConditions1}
\end{equation}
and
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}\tilde{w}_{i}\,c_{i}
  =\sum_{i=1}^{s}w_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}w_{i}\,c_{i}-\alpha=\f{1}{2}, 
  \label{orderConditions2}
\end{equation}
where $\tilde{c}_{i}$ and $c_{i}$ are given in \ref{app:butcherTables}.  
For globally stiffly accurate (GSA) IMEX schemes, $\tilde{w}_{i}=\tilde{a}_{si}$ and $w_{i}=a_{si}$ for $i=1,\ldots,s$, so that $\tilde{\vect{u}}^{n+1}=\vect{u}^{(s)}$ \cite{ascher_etal_1997}.  
This property is beneficial for very stiff problems and also simplifies the proof of the realizability-preserving property of the DG-IMEX scheme given in Section~\ref{sec:realizableDGIMEX}, since it eliminates the assembly step in Eq.~\eqref{imexIntermediate}.

Hu et al. \cite{hu_etal_2018} rewrite the stage values in Eq.~\eqref{imexStages} in the following form
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$ (see \ref{app:butcherTables}).  
In Eq.~\eqref{eq:imexStagesRewrite}, $\vect{u}^{(0)}=\vect{u}^{n}$.  
Two types of IMEX schemes are considered: \emph{type~A} \cite{pareschiRusso_2005,dimarcoPareschi2013} and \emph{type~ARS} \cite{ascher_etal_1997}.  
For IMEX schemes of type~A, the matrix $A$ is invertible.  
For IMEX schemes of type~ARS, the matrix $A$ can be written as
\begin{equation*}
  \left( 
    \begin{matrix} 
       0 & 0 \\ 
       0 & \hat{A}
    \end{matrix}
  \right),
\end{equation*}
where $\hat{A}$ is invertible.  
In writing the stages in the IMEX scheme in the general form given by Eq.~\eqref{eq:imexStagesRewrite}, it should be noted that $\tilde{c}_{i0}=0$ for IMEX schemes of Type~A, and $c_{i1}=\tilde{c}_{i1}=0$ for IMEX schemes of Type~ARS \cite{hu_etal_2018}.  
Type A schemes can be made convex-invariant by requiring 
\begin{align}
  &a_{ii}>0, \quad c_{i0}\ge0, \quad \text{for} \quad i=1,\ldots,s, \nonumber \\
  &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=2,\ldots,s, \quad\text{and}\quad j=1,\ldots,s-1.  
  \label{eq:positivityConditionsTypeA}
\end{align}
Similarly, type ARS schemes can be made convex-invariant by requiring 
\begin{align}
  &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
  &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1.  
  \label{eq:positivityConditionsTypeARS}
\end{align}
Coefficients were given in \cite{hu_etal_2018} for GSA schemes of type A with $s=3$ and type ARS with $s=4$.  
(It was also proven that $s=3$ and $s=4$ are the necessary number of stages needed for GSA second-order convex-invariant IMEX schemes of type A and type ARS, respectively.)

In Eq.~\eqref{eq:imexStagesRewrite}, the explicit part of the IMEX scheme has been written in the so-called Shu-Osher form \cite{shuOsher_1988}.  
For the scheme to be convex-invariant, the coefficients $c_{ij}$ must also satisfy $\sum_{j=0}^{i-1}c_{ij}=1$.  
Then, if the expression inside the square brackets in Eq.~\eqref{eq:imexStagesRewrite} --- which is in the form of a forward Euler update with time step $\hat{c}_{ij}\,\dt$ --- is in the (convex) set of admissible states for all $i=1,\ldots,s$, and $j=0,\ldots,i-1$, it follows from convexity arguments that the entire sum on the right-hand side of Eq.~\eqref{eq:imexStagesRewrite} is also admissible.  
Thus, if the explicit update with the transport operator is admissible for a time step $\dt_{\mbox{\tiny Ex}}$, the IMEX scheme is convex-invariant for a time step $\dt\le c_{\Sch}\,\dt_{\mbox{\tiny Ex}}$, where
\begin{equation}
  c_{\Sch}=\min_{\substack{i = 2,\ldots,s \\ 
       j = 1,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}} \quad \text{(Type A)}, \quad  c_{\Sch}=\min_{\substack{i = 2,\ldots,s \\ 
              j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}} \quad \text{(Type ARS)}.
  \label{eq:imexCFL}
\end{equation}
Here, $c_{\Sch}$ is the CFL condition, relative to $\dt_{\mbox{\tiny Ex}}$, for the IMEX scheme to be convex-invariant.  
It is desirable to make $c_{\Sch}$ as large (close to $1$) as possible.  
Note that for $\vect{u}^{(i)}$ to be admissible also requires the implicit solve (equivalent to implicit Euler) to be convex-invariant.  
In \cite{hu_etal_2018}, Hu et al. provide examples of GSA, convex-invariant IMEX schemes of type A (see scheme PA2 in \ref{app:butcherTables}) and type ARS.  
In \ref{app:butcherTables}, we provide another example of a GSA, convex-invariant IMEX scheme of type A (scheme PA2+), with a larger $c_{\Sch}$ (a factor of about $1.7$ larger).

\subsection{Convex-Invariant, Diffusion Accurate IMEX Schemes}

Unfortunately, the correction step in Eq.~\eqref{eq:imexCorrection} deteriorates the accuracy of the IMEX scheme when applied to the moment equations in the diffusion limit.  
The diffusion limit is characterized by frequent collisions ($\tau\ll 1$) in a purely scattering medium ($\xi=0$), and exhibits long-time behavior governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
Here the time derivative term in the equation for the particle flux has been dropped (formally $\cO(\tau^{2})$) so that, to leading order in $\tau$, the second equation in Eq.~\eqref{eq:diffusionLimit} states a balance between the transport term and the collision term.  
Furthermore, in the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$.  
The absence of the transport operator in the correction step in Eq.~\eqref{eq:imexCorrection}, destroys the balance between the transport term and the collision term.  
We demonstrate the inferior performance of IMEX schemes with this correction step in the diffusion limit in Section~\ref{sec:smoothProblems}.  
We have also implemented and tested one of the IMEX schemes in Chertock et al. \cite{chertock_etal_2015} (not included in Section~\ref{sec:smoothProblems}), where the transport operator is part of the correction step, and found it to perform very well in the diffusion limit.  
However, we have not been able to prove the realizability-preserving property with this approach without invoking a too severe time step restriction.  
We therefore proceed to design convex-invariant IMEX schemes without the correction step that perform well in the diffusion limit.  
We limit the scope to IMEX schemes of type~ARS.  
(It can be shown that IMEX schemes of type~A conforming to Definition~\ref{def:PD-IMEX} below do not exist; cf. \ref{app:noTypeA}.)

We take a heuristic approach to determine conditions on the coefficients of the IMEX scheme to ensure it performs well in the diffusion limit.  
We leave the spatial discretization unspecified.  
Define the vectors 
\begin{equation}
  \vec{\cJ}=(\cJ^{(1)},\ldots,\cJ^{(s)})^{T}
  \quad\text{and}\quad
  \vec{\vect{\cH}}=(\vect{\cH}^{(1)},\ldots,\vect{\cH}^{(s)})^{T}.  
\end{equation}
(The components of $\vec{\cJ}$ and $\vec{\vect{\cH}}$ can, e.g., be the cell averages evolved with the DG method.)  
We can then write the stages of the IMEX scheme in Eq.~\eqref{imexStages} applied to the particle density equation as
\begin{equation}
  \vec{\cJ}=\cJ^{n}\,\vec{\vect{e}} - \dt\,\tilde{A}\,\nabla\cdot\vec{\vect{\cH}}, 
  \label{eq:numberDensityIMEX}
\end{equation}
where $\vec{\vect{e}}$ is a vector of length $s$ containing all ones, and the divergence operator acts individually on the components of $\vec{\vect{\cH}}$.  
Similarly, for the particle flux equation we have
\begin{equation}
  \vec{\vect{\cH}}=\vect{\cH}^{n}\,\vec{\vect{e}}-\dt\,\big(\,\f{1}{3}\tilde{A}\,\nabla\vec{\cJ}+\f{1}{\tau}\,A\,\vec{\vect{\cH}}\,\big).
  \label{eq:numberFluxIMEX}
\end{equation}
In the context of IMEX schemes, the diffusion limit (cf. the second equation in \eqref{eq:diffusionLimit}) implies that the relation $A\,\vec{\vect{\cH}}=-\f{1}{3}\,\tau\,\tilde{A}\,\nabla\vec{\cJ}$ should hold.  
Define the pseudoinverse of the implicit coefficient matrix for IMEX schemes of type~ARS as
\begin{equation*}
    A^{-1}
    =\left(
        \begin{matrix}
          0 & 0 \\ 
          0 & \hat{A}^{-1}
        \end{matrix}
      \right).
\end{equation*}
Then, for the stages $i=1,\ldots,s$, 
\begin{equation}
  \vect{\cH}^{(i)}=-\f{1}{3}\,\tau\,\vec{\vect{e}}_{i}^{T}A^{-1}\tilde{A}\,\vec{\vect{e}}\,\nabla\cJ^{n}+\cO(\dt\,\tau^{2}),
\end{equation}
where $\vec{\vect{e}}_{i}$ is the $i$th column of the $s\times s$ identity matrix and we have introduced the expansion $\vec{\cJ}=\cJ^{n}\vec{\vect{e}}+\cO(\dt\,\tau)$.  
For $\vect{\cH}^{(i)}$ to be accurate in the diffusion limit, we require that
\begin{equation}
  \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s.
  \label{eq:diffusionCondition}
\end{equation}
(The case $i=1$ is trivial and does not place any constraints on the components of $A$ and $\tilde{A}$.)  
If Eq.~\eqref{eq:diffusionCondition} holds, then
\begin{equation}
  \f{\cJ^{(s)}-\cJ^{n}}{\dt}
  =-\tilde{\vect{w}}^{T}(\nabla\cdot\vec{\vect{\cH}})
  =\f{1}{3}\tau\,\nabla^{2}\cJ^{n}+\cO(\dt\,\tau^{2}), 
\end{equation}
which approximates a diffusion equation for $\cJ$ with the correct diffusion coefficient $\tau/3$.  
(For a GSA IMEX scheme without the correction step, $\cJ^{n+1}=\cJ^{(s)}$.)
